Finally Tagless, Partially Evaluated

### 2008-10-09 tagless/tagless-final.tex
take care of one more comment, and further adjust jfp-notes to reflect current status
Abstracting over an inductive map over object types Minor fixes also

Our principal technique is to encode
\ifshort HOAS \else de Bruijn or higher-order abstract syntax \fi
using combinator functions rather than data constructors.
In other words, we represent object terms not in an initial algebra
but using the coalgebraic structure of the $\lambda$-calculus.
Our representation also simulates inductive maps from types to
types, which are required for typed partial evaluation and CPS transformations.

Our encoding of an object term abstracts over the various ways to
interpret it, yet statically assures that the interpreters never get
stuck.  
%% To achieve self\hyp interpretation and show Jones\hyp
%% optimality, we relate this exemplar of higher-rank and higher-kind
%% polymorphism \ifshort\else (provided by ML functors and Haskell~98 
%% constructor classes) \fi to plugging a term into a context of 
%% let\hyp polymorphic bindings.

As a side-effect, we provide another convincing application of the 
usefulness of being able to abstract over higher-kinded types.
\end{abstract}

@@

We attack the problem of tagless (staged) type-preserving
interpretation exactly as it was posed
by \citet{WalidICFP02} and \citet{xi-guarded}.
We use their running examples and achieve the
result they call desirable.  Our contributions are as follows.
\begin{enumerate}
\item We build the first \emph{family} of interpreters, each instantiating the
   \emph{same} signature, that evaluate (\S\ref{language}),
   compile (\S\ref{S:compiler}), and 
   partially evaluate (\S\ref{PE}) a typed higher-order object language
   in a typed metalanguage, in direct and continuation\hyp passing
   styles\ifshort\else\ (\S\ref{variations})\fi.
\item These interpreters use no type tags
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
    Yet the type system of the metalanguage
    assures statically that each object program is well-typed and closed,
    and that each interpreter preserves types and never gets stuck.
    In particular, our (online) partial evaluator and CPS transformers
    avoid GADTs in their implementation and stay portable across
    Haskell 98 and ML, by expressing in their interface an inductive
    map from input types to output types.
\item Our clean, comparable implementations using OCaml modules and Haskell
    type classes show how to parametrize our final representation of object
    terms over multiple ways to assign them meanings.
\item We point a clear way to extend the object language with more features
    such as state\ifshort\else~(\S\ref{state})\fi.\ifshort\footnote{Again, please see our code.}\fi
    \ Our term encoding is contravariant in the object language, so
    extending the language does not invalidate terms already encoded.
%% \item We describe an approach to self\hyp interpretation compatible with the
%%   above\ifshort\else~(\S\ref{selfinterp})\fi.  Self\hyp interpretation turned
%%   out to be harder than expected.\ifshort\footnotemark[\value{footnote}]\fi
\item Show how higher-kinded abstraction can be useful for building
    embedded DSLs.
\end{enumerate}
Our code is surprisingly simple and obvious in hindsight, but
it has been cited as a difficult problem (\cite{sumii-hybrid}
and \cite{Thiemann-combinators}
notwithstanding) to interpret a typed object language in a typed metalanguage
without tagging or type\hyp system extensions.  For example, \citet{taha-tag}
say that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''  We discuss related work in~\S\ref{related}.

@@

Our approach accommodates
several variants, including
a call-by-name CPS interpreter and a call-by-value CPS
transformation.
\ifshort\label{state}\label{S:CPS}%
This lets us decouple the evaluation strategy of the object language
from that of the metalanguage. The accompanying code shows the CBN CPS
interpreter (module |RCN| implementing |Symantics|) and a CBV CPS
transformer |CPST|. The latter explicitly maps CPS interpretations to
(direct) interpretations performed by the base interpreter~|S|. All
these interpreters are typed, tagless and \emph{type-preserving} (as
well as fully polymorphic in the answer-type). The type preservation
is the consequence of the type soundness of the metalanguage.  We can
modify the CBV CPS transformation to pass a piece of state along with
the continuation. This technique lets us support mutable state.  Due
to the severe lack of space we cannot describe these interpreters and
refer the reader to the accompanying code.

@@

We would like to collect those interpretation functions into a module
with signature |Symantics|, to include the CBN CPS interpreter within our
general framework. Alas, as in~\S\ref{S:PE-problem}, the type of
an object term inductively determines the type of its interpretation:
the interpretation of an object term of type~$t$ may not have type
|(|$t$|->'w)->'w|, because $t$ may be a function type.  Again we
simulate a type function with a typecase distinction, using an extra
type argument to |repr|. Luckily, the type function |static| needed for
the partial evaluator 
in~\S\ref{S:PE-solution} is precisely the same type function we
need for CBN CPS\@.
\begin{code}
module RCN = struct
  type ('c,'sv,'dv) repr = {ko: 'w. ('sv -> 'w) -> 'w}
  let int (x:int) = {ko = fun k -> k x}
  let add e1 e2 = {ko = fun k ->
      e1.ko (fun v1 -> e2.ko (fun v2 -> k (v1 + v2)))}
  let if_ eb et ee = {ko = fun k ->
      eb.ko (fun vb -> if vb then (et ()).ko k else (ee ()).ko k)}
  let lam f = {ko = fun k -> k f}
  let app e1 e2 = {ko = fun k -> e1.ko (fun f -> (f e2).ko k)}
  let fix f = let rec fx f n = app (f (lam (fx f))) n in lam (fx f)
  let run x = x.ko (fun v -> v)
end
\end{code}

This interpreter~|RCN| is fully polymorphic over the answer type,
using higher-rank polymorphism through OCaml record types.
To avoid this higher-rank polymorphism in the core language, which
Standard ML does not support, we could also define |RCN| as
a functor parameterized over
the answer type.
\begin{code}
module RCN(W : sig type w end)= struct
  type ('c,'sv,'dv) repr = ('sv -> W.w) -> W.w
  ...
\end{code}
This allows a translation of these techniques to, for example, Standard ML\@.
Unfortunately, the loss of polymorphism in the answer type makes the 
CPS evaluator less pleasant to use than the other implementations.

@@

Changing one definition turns our CBN CPS interpreter into CBV\@.
\begin{code}
module RCV = struct include RCN
  let lam f = {ko = fun k -> k
      (fun e -> e.ko (fun v -> f {ko = fun k -> k v}))}
end
\end{code}
Now an applied abstraction
evaluates its argument before proceeding. This approach is in
line with \citets{reynolds-relation}, albeit typed. The
interpreter~|RCV| is useful for CBV evaluation of the object language
whether the metalanguage is CBV or CBN\@.

We turn to a more general approach to CBV CPS: a CPS transformer that
turns any implementation of |Symantics| into a CPS version of that
evaluator.
This functor on interpreters performs a textbook
CPS transformation on the object language.
\begin{code}
module CPST(S: Symantics) = struct
  let int i = S.lam (fun k -> S.app k (S.int i))
  let add e1 e2 = S.lam (fun k -> S.app e1 (S.lam (fun v1 ->
                                  S.app e2 (S.lam (fun v2 ->
                                  S.app k (S.add v1 v2))))))
  let lam f = S.lam (fun k -> S.app k
              (S.lam (fun x -> f (S.lam (fun k -> S.app k x)))))
  let app e1 e2 = S.lam (fun k -> S.app e1 (S.lam (fun f ->
                                  S.app e2 (S.lam (fun v ->
                                  S.app (S.app f v) k)))))
  let fix = S.fix
end
\end{code}
This (abbreviated) code explicitly maps CPS interpretations to
(direct) interpretations performed by 
the base interpreter~|S|.

The module returned by |CPST| does not define |repr|
and thus does not have signature |Symantics|.
The reason is again the type of |lam f|. Whereas
|int| and |add| return the (abbreviated) type
|('c, ..., (int -> 'w) -> 'w) S.repr|,
the type of \texttt{lam (add (int~1))} is
\begin{code}
('c, ..., ((int -> (int -> 'w1) -> 'w1) -> 'w2) -> 'w2) S.repr
\end{code}
Hence, to write the type equation defining |CPST.repr| we again need
a type function with a typecase distinction, similar to |static|
in~\S\ref{S:PE-solution}. Alas, the type function we need is not
identical to |static|, so we need to add another type argument to
|repr| in the |Symantics| signature. As in~\S\ref{S:PE-solution}, the
terms in previous implementations of |Symantics| stay unchanged, but the
|repr| type equations in those implementations have to take a new
(phantom) type argument.
The verbosity of these types is the only difficulty in defining a
new signature to replace |Symantics| that the output of |CPST| matches.

@@

@@

We can modify a CBN or CBV CPS transformation to pass a piece of state
along with the continuation. This technique lets us support mutable
state. As Figure~\ref{fig:state} shows, we extend our object language
with three imperative features.
\begin{enumerate}
    \item ``$\deref$'' gets the current state;
    \item ``$\set e$'' sets the state to the value of~$e$ and returns
        the previous value of the state;
    \item the let-form ``$\lapp{e_1}{x}e_2$'' evaluates $e_1$
        before~$e_2$ even if $e_2$ does not use~$x$.
\end{enumerate}
If $x$ does not appear in~$e_2$, then ``$\lapp{e_1}{x}e_2$'' is same as
the more familiar sequencing form ``$e_1;e_2$''.
We can embed this extended object language into OCaml by extending the
|Symantics| signature in Figure~\ref{fig:ocaml}.
\begin{code}
module type SymSI = sig
  include Symantics
  type state
  type 'c states       (* static version of the state *)
  val lapp : (('c,'sa,'da) repr as 'x) -> ('x -> 'y)
             -> (('c,'sb,'db) repr as 'y)
  val deref : unit -> ('c, 'c states, state) repr
  val set   : (('c, 'c states, state) repr as 'x) -> 'x
end
\end{code}
In HOAS\@, we write the term ``$\lapp{e_1}{x}e_2$'' as |lapp e1 (fun x -> e2)|;
the type of |lapp| is that of function application with
the two arguments swapped.  We can encode the term
``$\lapp{\deref}{x} (\set 2;\, x+\deref)$''
as the OCaml functor
\begin{code}
module EXSI_INT(S: SymSI
  with type state = int and type 'c states = int) = struct open S
  let test1 () = lapp (deref ()) (fun x -> 
                  lapp (set (int 2)) (fun _ -> add x (deref ())))
end
\end{code}
The accompanying source code shows several more tests, including
a test for higher-order state and a power function that uses state
as the accumulator.

The state-passing interpreter extends the CBN CPS
interpreter |RCN| of~\S\ref{S:CPS}.
\begin{code}
module RCPS(ST: sig 
  type state 
  type 'c states 
  type ('c,'sv,'dv) repr = 
      {ko: 'w. ('sv -> 'c states -> 'w) -> 'c states -> 'w}
end) = struct
  include ST
  type ('c, 'sv, 'dv) result = 'c states -> 'sv
  ... 
  let lapp e2 e1 = {ko = fun k ->
      e2.ko (fun v -> (app (lam e1) {ko = fun k -> k v}).ko k)}
  let deref () = {ko = fun k s -> k s s}
  let set e = {ko = fun k -> e.ko (fun v s -> k s v)}
  let get_res x = fun s0 -> x.ko (fun v s -> v) s0
end
\end{code}
The implementations of |int|, |app|, |lam|, and so on are
\emph{identical} to those of |RCN| and elided. New are the extended type
|repr|,
which now includes the state, and the functions
|lapp|, |deref|, and |set| representing imperative features. The
interpreter is still CBN, so evaluating |app ef ea| might not
evaluate |ea|, but evaluating |lapp ea ef| always does.
For first-order state, such as of type~$\ZZ$, we 
instantiate the interpreter as
\begin{code}
module RCPSI = RCPS(struct 
  type state = int
  type 'c states = int
  type ('c,'sv,'dv) repr = 
      {ko: 'w. ('sv -> 'c states -> 'w) -> 'c states -> 'w}
end)
\end{code}
If the state has a higher-order type, then the types |state| and |'c states|
are no longer the same, and |'s states| is mutually
recursive with the type |('c,'sv,'dv) repr|, as demonstrated in the
accompanying source code.

Because the |SymSI| signature extends |Symantics|, any encoding of
a term in the pure object language (that is, any functor that takes
a |Symantics| module as argument) can also be used as a term in the
extended object language (for example, applied to an implementation of
|SymSI|).  In particular,
|RCPSI| matches the |Symantics| signature and implements the
unextended object language: we can pass |RCPSI|
to the functor |EX| (Figure~\ref{fig:ocaml-simple}) and run 
the example |test1| from there. The main use for |RCPSI| is to interpret the 
extended object language.
\begin{code}
module EXPSI_INT = EXSI_INT(RCPSI)
let cpsitesti1 = RCPSI.get_res (EXPSI_INT.test1 ()) 100
val cpsitesti1 : int = 102
\end{code}

We reiterate that this implementation adding state and imperative
features is very close to the CPS interpreter and uses no new
techniques.
We can also add mutable references to the object language using
mutable references of the metalanguage, as shown in the accompanying
code.
\fi
\fi

\ifshort\else
\begin{comment}
\jacques{Should we really eliminate this entirely?}
\section{Self-interpretation}\label{selfinterp}

@@

Our separation between the |Symantics| interface and its many
implementations codifies the common practice of implementing an embedded
DSL by specifying an abstract syntax of object\hyp language pervasives,
such as addition and application, then providing multiple
interpretations of them.
The techniques we use to form such a family of interpreters
find their seeds in \citearound{'s \emph{language
triplets}}\citet{Holst-AMIX}, though in an untyped setting.
\Citet{JonesNielson:94:AbstractInterpretation} also
prefigured this separation when they decomposed a denotational definition of an
untyped object language into a core semantics (which we call abstract syntax)
and multiple interpretations.  When implementing an embedded DSL in this
way, it is also common practice to use phantom types to rule out
ill-typed object terms, as done in Lava
\citep{Lava} and by \citet{Rhiger-thesis}. However, these two approaches
are not tagless because they still use universal types, such as Lava's
\texttt{Bit} and \texttt{NumSig}, and Rhiger's \texttt{Raw} (his Figure~2.2)
and \texttt{Term} (his Chap.~3), which incur the attendant overhead of
pattern matching.  The universal type also
greatly complicates the soundness and completeness proofs of embedding
\citep{Rhiger-thesis}, whereas our proofs are trivial.
Rhiger's approach does not support typed CPS transformation (his~\S3.3.4).
\begin{comment}
Rhiger's But Fig 2.2, p33: universal type Raw.  He uses phantom type
upon the Exp datatype. But that is cheating: phantom type means
essentially we can easily do coerce. We use real types.  That's why he
had to do tedious proofs in Sec 2 of soundness and completeness of
embedding. Whereas our proofs are obvious.  His sec 3 is based on data
representation of terms. They have type tags.  We do nothing of that
kind: See Sec 3.1.2. See numerous "data Term" in Sec3, which is the U
type.  In Sec 3.3.4 (p76) Rhiger specifically says that his encoding
cannot do typed CPS transformation -- whereas our does. BTW, Rhiger
thesis contains the definitions of the interpreter and the compiler,
in the beginning. Use this in response to Rev1
\end{comment}

>>>>

Our principal technique is to encode
\ifshort HOAS \else de Bruijn or higher-order abstract syntax \fi
using combinator functions rather than data constructors.
In other words, we represent object terms not in an initial algebra
but using the coalgebraic structure of the $\lambda$-calculus.
Our representation also simulates inductive maps from types to
types, which are required for typed partial evaluation and CPS transformations.
Our encoding of an object term abstracts uniformly over the family of ways to
interpret it, yet statically assures that the interpreters never get
stuck.  
%% To achieve self\hyp interpretation and show Jones\hyp
%% optimality, we relate this exemplar of higher-rank and higher-kind
%% polymorphism \ifshort\else (provided by ML functors and Haskell~98 
%% constructor classes) \fi to plugging a term into a context of 
%% let\hyp polymorphic bindings.
This family of interpreters thus demonstrates again that it is
useful to abstract over higher-kinded types.
\end{abstract}

@@

We attack the problem of tagless (staged) type-preserving
interpretation exactly as it was posed
by \citet{WalidICFP02} and \citet{xi-guarded}.
We use their running examples and achieve the
result they call desirable.  Our contributions are as follows.
\begin{enumerate}
\item We build the first \emph{family} of interpreters, each instantiating the
   \emph{same} signature, that evaluate (\S\ref{language}),
   compile (\S\ref{S:compiler}), and 
   partially evaluate (\S\ref{PE}) a typed higher-order object language
   in a typed metalanguage, in direct and continuation\hyp passing
   styles\ifshort\else\ (\S\ref{variations})\fi.
\item These interpreters use no type tags
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
    Yet the type system of the metalanguage
    assures statically that each object program is well-typed and closed,
    and that each interpreter preserves types and never gets stuck.
    In particular, our (online) partial evaluator and CPS transformers
    avoid GADTs in their implementation and stay portable across
    Haskell 98 and ML, by expressing in their interface an inductive
    map from input types to output types.
\item Our clean, comparable implementations using OCaml modules and Haskell
    type classes show how to parametrize our final representation of object
    terms over multiple ways to assign them meanings.
\item We point a clear way to extend the object language with more features
    such as state\ifshort\else~(\S\ref{state})\fi.\ifshort\footnote{Again, please see our code.}\fi
    \ Our term encoding is contravariant in the object language, so
    extending the language does not invalidate terms already encoded.
%% \item We describe an approach to self\hyp interpretation compatible with the
%%   above\ifshort\else~(\S\ref{selfinterp})\fi.  Self\hyp interpretation turned
%%   out to be harder than expected.\ifshort\footnotemark[\value{footnote}]\fi
\item We show how to use higher-kinded abstraction to build
    embedded DSLs.
\end{enumerate}
Our code is surprisingly simple and obvious in hindsight, but
it has been cited as a difficult problem (\cite{sumii-hybrid}
and \cite{Thiemann-combinators}
notwithstanding) to interpret a typed object language in a typed metalanguage
without tagging or type\hyp system extensions.  For example, \citet{taha-tag}
say that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''  We discuss related work in~\S\ref{related}.

@@

Our approach accommodates
several variants, including
a call-by-name CPS interpreter and a call-by-value CPS
transformation.
Of course, CPS is a well-studied topic, and \citearound{'s work on
program generation}\citet{Thiemann-combinators} already includes a
CPS evaluator expressed
using combinator functions rather than data constructors.
We focus here on expressing CPS transformations as part of a larger,
typed family of interpreters.
\ifshort\label{state}\label{S:CPS}%
This lets us decouple the evaluation strategy of the object language
from that of the metalanguage. The accompanying code shows the CBN CPS
interpreter (module |RCN| implementing |Symantics|) and a CBV CPS
transformer |CPST|. The latter explicitly maps CPS interpretations to
(direct) interpretations performed by the base interpreter~|S|. All
these interpreters are typed, tagless and \emph{type-preserving} (as
well as fully polymorphic in the answer-type). The type preservation
is the consequence of the type soundness of the metalanguage.  We can
modify the CBV CPS transformation to pass a piece of state along with
the continuation. This technique lets us support mutable state.  Due
to the severe lack of space we cannot describe these interpreters and
refer the reader to the accompanying code.

@@

We would like to collect those interpretation functions into a module
with signature |Symantics|, to include the CBN CPS interpreter within our
general framework. Alas, as in~\S\ref{S:PE-problem}, the type of
an object term inductively determines the type of its interpretation:
the interpretation of an object term of type~$t$ may not have type
|(|$t$|->'w)->'w|, because $t$ may be a function type.  Again we
simulate a type function with a typecase distinction, by changing the
type arguments to |repr|. Luckily, the type function |static| needed for
the partial evaluator 
in~\S\ref{S:PE-solution} is precisely the same type function we
need for CBN CPS\@, so our CBN interpreter can match the
|Symantics| signature in~\S\ref{S:PE-solution}, without even
using the |'dv| argument to |repr|.
\begin{code}
module RCN = struct
  type ('c,'sv,'dv) repr = {ko: 'w. ('sv -> 'w) -> 'w}
  let int (x:int) = {ko = fun k -> k x}
  let add e1 e2 = {ko = fun k ->
      e1.ko (fun v1 -> e2.ko (fun v2 -> k (v1 + v2)))}
  let if_ eb et ee = {ko = fun k ->
      eb.ko (fun vb -> if vb then (et ()).ko k else (ee ()).ko k)}
  let lam f = {ko = fun k -> k f}
  let app e1 e2 = {ko = fun k -> e1.ko (fun f -> (f e2).ko k)}
  let fix f = let rec fx f n = app (f (lam (fx f))) n in lam (fx f)
  let run x = x.ko (fun v -> v)
end
\end{code}

This interpreter~|RCN| is fully polymorphic over the answer type,
using higher-rank polymorphism through OCaml record types.
To avoid this higher-rank polymorphism in the core language,
we could also define |RCN| as
a functor parameterized over
the answer type.
\begin{code}
module RCN(W: sig type w end) = struct
  type ('c,'sv,'dv) repr = ('sv -> W.w) -> W.w   ...
\end{code}
This alternative is more cumbersome to use because the functor
needs to applied once for each answer type, but it translates
to, for example, Standard ML\@, whose core language does not
support higher-rank polymorphism.

@@

Changing one definition turns our CBN CPS interpreter into CBV\@.
\begin{code}
module RCV = struct include RCN
  let lam f = {ko = fun k -> k
      (fun e -> e.ko (fun v -> f {ko = fun k -> k v}))}
end
\end{code}
Now an applied abstraction
evaluates its argument before proceeding.  The
interpreter~|RCV| is useful for CBV evaluation of the object language
whether the metalanguage is CBV or CBN\@.
To match the same |Symantics| signature as |RCN| above, |RCV| uses
\citets{reynolds-relation} CBV CPS transformation, in which variables
denote computations (that is, functions from continuations),
rather than \citets{PlotkinCBN}, in which variables denote values.

We turn to a more general approach to CBV CPS: a CPS transformer that
turns any implementation of |Symantics| into a CPS version of that
evaluator.
This functor on interpreters performs \citets{PlotkinCBN} textbook
CPS transformation on the object language.
\begin{code}
module CPST(S: Symantics) = struct
  let int i = S.lam (fun k -> S.app k (S.int i))
  let add e1 e2 = S.lam (fun k -> S.app e1 (S.lam (fun v1 ->
                                  S.app e2 (S.lam (fun v2 ->
                                  S.app k (S.add v1 v2))))))
  let lam f = S.lam (fun k -> S.app k
              (S.lam (fun x -> f (S.lam (fun k -> S.app k x)))))
  let app e1 e2 = S.lam (fun k -> S.app e1 (S.lam (fun f ->
                                  S.app e2 (S.lam (fun v ->
                                  S.app (S.app f v) k)))))
  let fix = S.fix
end
\end{code}
This (abbreviated) code explicitly maps CPS interpretations to
(direct) interpretations performed by 
the base interpreter~|S|.

The module returned by |CPST| does not define |repr|
and thus does not have signature |Symantics|.
The reason is again the type of |lam f|. Whereas
|int| and |add| return the (abbreviated) type
|('c, ..., (int -> 'w) -> 'w) S.repr|,
the type of \texttt{lam (add (int~1))} is
\begin{code}
('c, ..., ((int -> (int -> 'w1) -> 'w1) -> 'w2) -> 'w2) S.repr
\end{code}
Hence, to write the type equation defining |CPST.repr| we again need
a type function with a typecase distinction, similar to |static|
in~\S\ref{S:PE-solution}. Alas, the type function we need is not
identical to |static|, so again we need to change the type arguments to
|repr| in the |Symantics| signature. As in~\S\ref{S:PE-solution}, the
terms in previous implementations of |Symantics| stay unchanged, but the
|repr| type equations in those implementations have to take a new
(phantom) type argument.
The verbosity of these types is the only difficulty in defining a
new signature to replace |Symantics| that the output of |CPST| matches.

@@

\subsection{Abstracting over an inductive map on object types}

\begin{figure}[t]
\begin{floatrule}
\begin{code}
module type Symantics1 = sig
  type 'c dint
  type 'c dbool
  type ('c,'da,'db) darr
  type ('c,'dv) repr
  val int : int  -> ('c, 'c dint) repr
  val bool: bool -> ('c, 'c dbool) repr
  val lam : (('c,'da) repr -> ('c,'db) repr) -> ('c, ('c,'da,'db) darr) repr
  val app : ('c, ('c,'da,'db) darr) repr -> ('c, 'da) repr -> ('c, 'db) repr
  val fix : ('x -> 'x) -> (('c, ('c,'da,'db) darr) repr as 'x)
  val add : ('c, 'c dint) repr -> ('c, 'c dint) repr -> ('c, 'c dint) repr
  val mul : ('c, 'c dint) repr -> ('c, 'c dint) repr -> ('c, 'c dint) repr
  val leq : ('c, 'c dint) repr -> ('c, 'c dint) repr -> ('c, 'c dbool) repr
  val if_ : ('c, 'c dbool) repr
            -> (unit -> 'x) -> (unit -> 'x) -> (('c, 'da) repr as 'x)
end
\end{code}
\end{floatrule}
\caption{A (Meta)OCaml embedding that abstracts over an inductive map on object types}
\label{fig:inductive}
\end{figure}

Having seen that each CPS interpreter above matches a differently
modified |Symantics| signature, one may wonder whether |Symantics|
can be generalized to encompass them all.  The answer is yes: the
|Symantics1| signature in Figure~\ref{fig:inductive} abstracts our
representation of object terms not only over the type constructor
|repr| but also over the three branches that make up an inductive
map, such as |static| in~\S\ref{S:PE-solution}, from object types
to metalanguage types.  The first two branches (for the object
types $\ZZ$ and~$\BB$) become the abstract types |dint| and
|dbool|, whereas the third branch (for object types $t_1\to t_2$)
becomes the new abstract type constructor |darr|.

Almost every interpreter in this paper can be made to match the
|Symantics1| signature without changing any terms, by defining
|dint|, |dbool|, |darr|, and |repr| suitably.  For example,
the types in the evaluator~|R| and the CBV CPS transformer |CPST|
should be changed as follows.
\begin{code}[commandchars=\@\[\]]
module R = struct
  type 'c dint = int
  type 'c dbool = bool
  type ('c,'da,'db) darr = 'da -> 'db
  type ('c,'dv) repr = 'dv   ...

module CPST(S: Symantics1)(W: sig type 'c dw end) = struct open W
  type 'c dint = 'c S.dint
  type 'c dbool = 'c S.dbool
  type ('c,'da,'db) darr =
    ('c, 'da, ('c, ('c, 'db, 'c dw) S.darr, 'c dw) S.darr) S.darr
  type ('c,'dv) repr =
    ('c, ('c, ('c, 'dv, 'c dw) S.darr, 'c dw) S.darr) S.repr   ...
\end{code}
Modified thus, |CPST| produces modules that match |Symantics1| and
thus can be not only evaluated or compiled but also transformed
using |CPST| again.  The accompanying source file |inco.ml| shows
the details, including one-pass CPS transformations in the
higher-order style of \citet{danvy-representing}.

The only interpreter in this paper that does not fit |Symantics1|
is the partial evaluator~|P|.  It does not fit because it uses
\emph{two} inductive maps on object types---both |'sv| and |'dv|
in Figure~\ref{fig:ocaml}.  We could define a |Symantics2|
signature to abstract over \emph{two} inductive maps over object
types; it would include 4 abstract types and 2 abstract type
constructors in addition to |repr|.  It would then be easy to
make~|P| match |Symantics2|, but the output of~|P| can still only
match |Symantics1|.  This escalation points to a need for either
record-kind polymorphism (so that |'dv| in
Figure~\ref{fig:inductive} may be more than just one type) or
type-indexed types (so that we do not need to emulate them in the
first place).

@@

We can modify a CBN or CBV CPS transformation to pass a piece of state
along with the continuation. This technique lets us support mutable
state (or more generally any monadic effect) by representing it
using continuations \citep{filinski-representing}.
As Figure~\ref{fig:state} shows, we extend our object language
with three imperative features.
\begin{enumerate}
    \item ``$\deref$'' gets the current state;
    \item ``$\set e$'' sets the state to the value of~$e$ and returns
        the previous value of the state;
    \item the let-form ``$\lapp{e_1}{x}e_2$'' evaluates $e_1$
        before~$e_2$ even if $e_2$ does not use~$x$ and even if
        evaluation is CBN\@.
\end{enumerate}
The form ``$\lapp{e_1}{x}e_2$'' is equivalent to
``let'' in \citearound{'s monadic metalanguage}\citet{moggi-notions}.
If $x$ does not appear in~$e_2$, then it is same as
the more familiar sequencing form ``$e_1;e_2$''.
We embed this extended object language into OCaml by extending the
|Symantics| signature in Figure~\ref{fig:ocaml}.
\begin{code}
module type SymSI = sig
  include Symantics
  type state
  type 'c states       (* static version of the state *)
  val lapp : (('c,'sa,'da) repr as 'x) -> ('x -> 'y)
             -> (('c,'sb,'db) repr as 'y)
  val deref : unit -> ('c, 'c states, state) repr
  val set   : (('c, 'c states, state) repr as 'x) -> 'x
end
\end{code}
In HOAS\@, we write the term ``$\lapp{e_1}{x}e_2$'' as |lapp e1 (fun x -> e2)|;
the type of |lapp| is that of function application with
the two arguments swapped.  We can encode the term
``$\lapp{\deref}{x} (\set 2;\, x+\deref)$''
as the OCaml functor
\begin{code}
module EXSI_INT(S: SymSI
  with type state = int and type 'c states = int) = struct open S
  let test1 () = lapp (deref ()) (fun x -> 
                  lapp (set (int 2)) (fun _ -> add x (deref ())))
end
\end{code}
The accompanying source code shows several more tests, including
a test for higher-order state and a power function that uses state
as the accumulator.

The state-passing interpreter extends the CBN CPS
interpreter |RCN| of~\S\ref{S:CPS}.
\begin{code}
module RCPS(ST: sig 
  type state 
  type 'c states 
  type ('c,'sv,'dv) repr = 
      {ko: 'w. ('sv -> 'c states -> 'w) -> 'c states -> 'w}
end) = struct include ST ... 
  let lapp e2 e1 = {ko = fun k ->
      e2.ko (fun v -> (app (lam e1) {ko = fun k -> k v}).ko k)}
  let deref () = {ko = fun k s -> k s s}
  let set e = {ko = fun k -> e.ko (fun v s -> k s v)}
  let get_res x = fun s0 -> x.ko (fun v s -> v) s0
end
\end{code}
The implementations of |int|, |app|, |lam|, and so on are
\emph{identical} to those of |RCN| and elided. New are the extended type
|repr|,
which now includes the state, and the functions
|lapp|, |deref|, and |set| representing imperative features. The
interpreter is still CBN, so evaluating |app ef ea| might not
evaluate |ea|, but evaluating |lapp ea ef| always does.
For first-order state, such as of type~$\ZZ$, we 
instantiate the interpreter as
\begin{code}
module RCPSI = RCPS(struct 
  type state = int
  type 'c states = int
  type ('c,'sv,'dv) repr = 
      {ko: 'w. ('sv -> 'c states -> 'w) -> 'c states -> 'w}
end)
\end{code}
If the state has a higher-order type, then the types |state| and |'c states|
are no longer the same, and |'s states| is mutually
recursive with the type |('c,'sv,'dv) repr|, as demonstrated in the
accompanying source code.

Because the |SymSI| signature extends |Symantics|, any encoding of
a term in the pure object language (that is, any functor that takes
a |Symantics| module as argument) can also be used as a term in the
extended object language (for example, applied to an implementation of
|SymSI|).  In particular,
|RCPSI| matches the |Symantics| signature and implements the
unextended object language: we can pass |RCPSI|
to the functor |EX| (Figure~\ref{fig:ocaml-simple}) and run 
the example |test1| from there. The main use for |RCPSI| is to interpret the 
extended language.
\begin{code}
module EXPSI_INT = EXSI_INT(RCPSI)
let cpsitesti1 = RCPSI.get_res (EXPSI_INT.test1 ()) 100
val cpsitesti1 : int = 102
\end{code}

We reiterate that this implementation adding state and imperative
features is very close to the CPS interpreter and uses no new
techniques.
We can also add mutable references to the object language using
mutable references of the metalanguage, as shown in the accompanying
code.
Yet another way to add side effects to the object language is to
write a monadic interpreter (for a specific monad or a general
class of monads), which can be structured as a module matching the
|Symantics1| signature in Figure~\ref{fig:inductive}.
\fi
\fi

\ifshort\else
\begin{comment}
\jacques{Should we really eliminate this entirely?}
\section{Polymorphism over interpreters}

It is crucial to our achievements above that the same object term can be
interpreted by multiple interpreters, whether an evaluator, compiler, partial
evaluator, or CPS transformer.  In the typical \emph{initial} approach to DSL
embedding, interpretation is represented in the metalanguage by applying the
interpreter to the object term.  Flipping this application around, our
\emph{final} approach applies the object term to the interpreter.  In this
section, we explain how this flip amounts to Church-encoding the recursive GADT
of well-typed object terms and how it affects the notion of self-interpretation
and Jones optimality.

At the type level, an interpreter is represented by its type constructor
|repr|.  For example, a |Symantics| module or instance that matches the simple
encoding in~\S\ref{language} is in System~$F_\omega$ an interpreter

, over which an object term is a polymorphic function---either an ML
functor from a |Symantics| module containing |repr| or a Haskell value with
a |Symantics| constraint over |repr|.

\section{Self-interpretation}\label{selfinterp}

@@

Our separation between the |Symantics| interface and its many
implementations codifies the common practice of implementing an embedded
DSL by specifying an abstract syntax of object\hyp language pervasives,
such as addition and application, then providing multiple
interpretations of them.
The techniques we use to form such a family of interpreters
find their seeds in \citearound{'s \emph{language
triplets}}\citet{Holst-AMIX}, though in an untyped setting.  In the typed
setting, \citet{Nielson88} uses a typed $\lambda$-calculus as a metalanguage
for expressing families of program analyses; however the embedding of the
object language into the metalanguage, unlike our |Symantics| is not typed.
\Citet{JonesNielson:94:AbstractInterpretation} also
prefigured this separation when they decomposed a denotational definition of an
untyped object language into a core semantics (which we call abstract syntax)
and multiple interpretations.  When implementing an embedded DSL in this
way, it is also common practice to use phantom types to rule out
ill-typed object terms, as done in Lava
\citep{Lava} and by \citet{Rhiger-thesis}. However, these two approaches
are not tagless because they still use universal types, such as Lava's
\texttt{Bit} and \texttt{NumSig}, and Rhiger's \texttt{Raw} (his Figure~2.2)
and \texttt{Term} (his Chap.~3), which incur the attendant overhead of
pattern matching.  The universal type also
greatly complicates the soundness and completeness proofs of embedding
\citep{Rhiger-thesis}, whereas our proofs are trivial.
Rhiger's approach does not support typed CPS transformation (his~\S3.3.4).
\begin{comment}
Rhiger's But Fig 2.2, p33: universal type Raw.  He uses phantom type
upon the Exp datatype. But that is cheating: phantom type means
essentially we can easily do coerce. We use real types.  That's why he
had to do tedious proofs in Sec 2 of soundness and completeness of
embedding. Whereas our proofs are obvious.  His sec 3 is based on data
representation of terms. They have type tags.  We do nothing of that
kind: See Sec 3.1.2. See numerous "data Term" in Sec3, which is the U
type.  In Sec 3.3.4 (p76) Rhiger specifically says that his encoding
cannot do typed CPS transformation -- whereas our does. BTW, Rhiger
thesis contains the definitions of the interpreter and the compiler,
in the beginning. Use this in response to Rev1
\end{comment}
